Bare Demo of IEEEtran.cls\\ for IEEE Conferences

### 2016-11-30 paper.tex
Correcting typo with naïve Bayes

Learning to program is often regarded as a difficult task, courses are 
problem-solving intensive and often students feel frustrated when doing programming
exercises. Instructors aim to design learning activities that result
challenging to students, but not to the extreme as to frustrate them. When
selecting an appropiate programing excercise, experienced instructors gauge a
student’s affective state and skills to then assign an activity with the  appropriate level
of difficulty. This work is focused on the prediction of the affective states
of programmers with different levels of expertise when learning a new
programming language. For this, an interactive web-based programming platform
is proposed. The platform is designed to collect data from the students'
interaction for data analysis. Current work is focused on the prediction of
affective states using non-obtrusive sensors. Specifically, the proposed method uses data
gathered from both keyboard and mouse dynamics. The data set was generated
as students learned the basics of Python programming. Feature vectors 
were generated with 39 features representing the dynamics of the interaction. 
In total 149 feature vectors were generated and used as input of four different
classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
Networks. These classification models are then compared by their performances
when classifying learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Each of the classification
models achieved an accuracy of around 75\% with a kappa of around 0.45, and, in
average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
that the data gathered from the non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states.

@@

%But the problem is how to recognize this state so that it can be used
%to (select appropriate content| give grades| whatever ) - JJ
In this work, a method for the recognition of affective states through keystroke
and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
of keystrokes according to their affective state when programming. 
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
 The method
has three main components: a Javascript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using the web based learning environment. Using only the data from the
student’s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a naïve Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise. The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task. In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.

@@

Once feature vectors are obtained from an experiment, the generated dataset is
normally used as training data for a classifier. Researchers of affective
recognition have used a wide variety of classification algorithms. As a proof of
concept four well known classification algorithms where compared: k-Nearest
Neighbors (k-NN) algorithm, a feed forward neural network trained with back-
propagation, a naïve Bayes classifier and finally a decision trees algorithm for
rational data (J-48). Details for these algorithms, can be found in the textbook
by Tan, Steinbach, \& Kumar \cite{tan2006introduction}. The accuracies obtained by these algorithms
as well as the parameters used are reported in the next section.

@@

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network, but the fact that they needed Facebook kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services; however, the use of Facebook gave researchers access to public
information about the subjects:  all the learners were either current students or
graduates of Computer Systems Engineering. Their ages were in the range of 18 to
30 years. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, all the participants were
required had completed at least one course in software programming. Results from
the initial survey are shown in Figure 4. %Please use reference. Fig 4
                                %is completed activities. I think this
                                %should be Fig 5. 
\begin{table*}[!t]
\centering
\caption{The performance of the classifiers.}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Naïve Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.11) 0.50   & 80.48\% (0.10) 0.51 & 76.14(0.09) 0.43    & 78.43(0.12) 0.45 \\
    \hline
    Relaxation      & 71.10\% (0.09) 0.40   & 72.57\% (0.11) 0.45 & 69.14\%(0.08)  0.37 & 71.24\%(0.07)  0.34\\
    \hline
    Distraction     & 70.33\% (0.13)  0.43  & 71.00\% (0.10) 0.43 & 74.00\%(0.06)  0.50 & 72.52\%(0.07)  0.39\\
    \hline
    Frustration     & 74.71\% (0.09) 0.49   & 73.86\% (0.12) 0.45 & 72.62\%(0.09)  0.42 & 78.0\%(0.09)  0.51 \\
    \hline
    Boredom         & 74.71\% (0.13) 0.47   & 84.57\% (0.06) 0.59 & 83.81\%(0.06)  0.58 & 76.19\%(0.09)  0.45\\
    \hline
    \end{tabular}
\end{table*}

@@

\section{Conclusion}
The aim of this research was to evaluate the use of keyboard and mouse dynamics 
as an appropriate sensory input for an affective recognition system. The context
of use would be learning environment for programmers. In particular for learning
activities consisting in writing short programs interactively. An experimental
approach was adopted with this aim. The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and kappa below their counterparts of fixed-text, and in this
work the results obtained were similar to those works applied to fixed-text 
dynamics. A possible explanation of this would be the addition of the mouse 
dynamics features and the additional preprocessing performed on the feature vectors. 
While these are promising results further experiments are needed. Other experiments
should focus on novice programmers.
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?

>>>>

Learning to program is often regarded as a difficult task, courses are 
problem-solving intensive and often students feel frustrated when doing programming
exercises. Instructors aim to design learning activities that result
challenging to students, but not to the extreme as to frustrate them. When
selecting an appropiate programing excercise, experienced instructors gauge a
student’s affective state and skills to then assign an activity with the  appropriate level
of difficulty. This work is focused on the prediction of the affective states
of programmers with different levels of expertise when learning a new
programming language. For this, an interactive web-based programming platform
is proposed. The platform is designed to collect data from the students'
interaction for data analysis. Current work is focused on the prediction of
affective states using non-obtrusive sensors. Specifically, the proposed method uses data
gathered from both keyboard and mouse dynamics. The data set was generated
as students learned the basics of Python programming. Feature vectors 
were generated with 39 features representing the dynamics of the interaction. 
In total 149 feature vectors were generated and used as input of four different
classification algorithms: Na\"ive Bayes, J-48, k-NN, and Artificial Neural
Networks. These classification models are then compared by their performances
when classifying learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Each of the classification
models achieved an accuracy of around 75\% with a kappa of around 0.45, and, in
average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
that the data gathered from the non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states.

@@

%But the problem is how to recognize this state so that it can be used
%to (select appropriate content| give grades| whatever ) - JJ
In this work, a method for the recognition of affective states through keystroke
and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
of keystrokes according to their affective state when programming. 
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
 The method
has three main components: a Javascript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using the web based learning environment. Using only the data from the
student’s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise. The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task. In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.

@@

Once feature vectors are obtained from an experiment, the generated dataset is
normally used as training data for a classifier. Researchers of affective
recognition have used a wide variety of classification algorithms. As a proof of
concept four well known classification algorithms where compared: k-Nearest
Neighbors (k-NN) algorithm, a feed forward neural network trained with back-
propagation, a na\"ive Bayes classifier and finally a decision trees algorithm for
rational data (J-48). Details for these algorithms, can be found in the textbook
by Tan, Steinbach, \& Kumar \cite{tan2006introduction}. The accuracies obtained by these algorithms
as well as the parameters used are reported in the next section.

@@

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network, but the fact that they needed Facebook kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services; however, the use of Facebook gave researchers access to public
information about the subjects:  all the learners were either current students or
graduates of Computer Systems Engineering. Their ages were in the range of 18 to
30 years. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, all the participants were
required had completed at least one course in software programming. Results from
the initial survey are shown in Figure 4. %Please use reference. Fig 4
                                %is completed activities. I think this
                                %should be Fig 5. 
\begin{table*}[!t]
\centering
\caption{The performance of the classifiers.}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.11) 0.50   & 80.48\% (0.10) 0.51 & 76.14(0.09) 0.43    & 78.43(0.12) 0.45 \\
    \hline
    Relaxation      & 71.10\% (0.09) 0.40   & 72.57\% (0.11) 0.45 & 69.14\%(0.08)  0.37 & 71.24\%(0.07)  0.34\\
    \hline
    Distraction     & 70.33\% (0.13)  0.43  & 71.00\% (0.10) 0.43 & 74.00\%(0.06)  0.50 & 72.52\%(0.07)  0.39\\
    \hline
    Frustration     & 74.71\% (0.09) 0.49   & 73.86\% (0.12) 0.45 & 72.62\%(0.09)  0.42 & 78.0\%(0.09)  0.51 \\
    \hline
    Boredom         & 74.71\% (0.13) 0.47   & 84.57\% (0.06) 0.59 & 83.81\%(0.06)  0.58 & 76.19\%(0.09)  0.45\\
    \hline
    \end{tabular}
\end{table*}

@@

\section{Conclusion}
The aim of this research was to evaluate the use of keyboard and mouse dynamics 
as an appropriate sensory input for an affective recognition system. The context
of use would be learning environment for programmers. In particular for learning
activities consisting in writing short programs interactively. An experimental
approach was adopted with this aim. The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and kappa below their counterparts of fixed-text, and in this
work the results obtained were similar to those works applied to fixed-text 
dynamics. A possible explanation of this would be the addition of the mouse 
dynamics features and the additional preprocessing performed on the feature vectors. 
While these are promising results further experiments are needed. Other experiments
should focus on novice programmers.
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective?
